Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC with TMM normlaisation)

### 2016-11-05 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
Added markdown to process DLPFC and ACC together with a random effects model

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'ACC - STAR Reprocessed - with TMM')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC)';

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```
Explore metatdata
```{r dataExplore, fig.width = 10, fig.height = 10}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'RIN';

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
```

### Normlaisation (Custom design)
Adding Dx, Institution, Gender, EV1, EV3, EV4 and EV5 to the list of postadjust covariates RIN, RIN2, Age of Death, PMI and EV2
```{r custom.design}
# Add primary variable to the list (in case not chosen by iterative method)
postAdjustCovars = c('Dx', c(FactorCovariates, ContCovariates)) %>% unique()

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', '),
                 'as fixed effects'))

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape = Institution, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p

# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxControl-DxOther",
                                     "DxControl-DxSCZ",
                                     "DxOther-DxSCZ"),
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM", parentId = parentId)
CODE <- synStore(CODE)

>>>>

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'ACC - STAR Reprocessed - with TMM')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC)';

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```

Correlation/association between covariates at an FDR <= 0.1

```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

Explore metatdata

```{r dataExplore, fig.width = 14, fig.height = 6}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=10, fig.height=6}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis)

Adding Instituion and Gender a priori to variable selection as they are correlated with Dx
```{r iterativeNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx', 'Institution', 'Gender');

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

### Sanity check
```{r residualSigCovars.manual, fig.width=10, fig.height=6}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape = Institution, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p
```

```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxControl-DxOther",
                                     "DxControl-DxSCZ",
                                     "DxOther-DxSCZ"),
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM", parentId = parentId)
CODE <- synStore(CODE)
